Practical Optional Types for Clojure

### 2015-02-25 overview.tex
HMap overview

\subsection{Multimethod}

A multimethod in Clojure is a function with a dispatch mechanism to
of internal methods. Multimethods are created with {\clj{defmulti}}
and methods are installed via {\clj{defmethod}}.
They are called like regular functions.

\begin{Code}
\begin{exmp}
\begin{minted}{clojure}
(ann rep [Any -> String])
(defmulti rep class)
(defmethod rep Number [x] (str (inc x)))
(defmethod rep Keyword [x] (str (name x)))

(rep :a) ;=> "a"
(rep 1)  ;=> "2"
\end{minted}
\end{exmp}
\end{Code}

The \clj{class} function returns the class of an object, or \clj{nil}
if given \clj{nil}. This multimethod's dispatch function returns the class of its
argument. The underlying dispatch strategy features \clj{isa?}, a function
that is a subclassing check when provided with classes

\begin{minted}{clojure}
  (isa? Keyword Object) ;=> true
  (isa? Keyword Number) ;=> false
\end{minted}

otherwise an equality check.

\begin{minted}{clojure}
  (isa? :a :a) ;=> true
  (isa? :a 1) ;=> false
\end{minted}

So in the previous example, \clj{(rep :a)} dispatched to the method
where \clj{(isa? (class :a) v)} returned true, where \clj{v} is the dispatch value.
Since \clj{(isa? (class :a) Keyword)} is true, the second method is chosen.

The flexibility of \clj{isa?} is key to the generality of multimethods. 
For instance, we can dispatch on the value of
a map entry, even if it is not a class.

\begin{Code}
\begin{exmp}
\inputminted[firstline=4,lastline=23]{clojure}{code/demo/src/demo/eg5.clj}
\end{exmp}
\end{Code}

\clj{isa?} is further special-case with vectors: with two vectors of the
same length, it returns the logical conjunction of the pairwise applications
of \clj{isa?}.

\begin{minted}{clojure}
  (isa? [Keyword :a] [Object :a]) ;=> true
  (isa? [Keyword Keyword] [Object Object]) ;=> true
\end{minted}

We can use this to dispatch on more than one thing. Here we perform
double dispatch on the class of each argument.

\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
\end{exmp}
\end{Code}

No extra annotations are needed to follow the type-directed control flow.
For instance, the middle method is has inferred the left argument is a number
and thus safe to increment, and the right argument is a keyword and safe to
extract its name.

A dispatch value can also be the keyword \clj{:default} to specify a default
method if no preferred method is found. The static type of both arguments
remains unchanged.

@@

The final example combines everything we will cover for the rest of the paper:
multimethod dispatch, reflection resolution via type hints, Java method
and constructor calls, conditional and exceptional flow reasoning
and use of HMaps.

\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
\end{exmp}
\end{Code}

>>>>

\subsection{Multimethods}

A multimethod in Clojure is a function that contains a dispatch
function and methods. Multimethods are created with {\clj{defmulti}}.
\begin{minted}{clojure}
(ann rep [Any -> String])
(defmulti rep class)
\end{minted}
\clj{rep} is a multimethod of type \clj{[Any -> String]} with an empty \emph{dispatch table}
and \emph{dispatch function} \clj{class}, a function that returns the argument's class or \clj{nil} if none.
There are no methods to dispatch to so invocations of \clj{rep} will fail at runtime.
\begin{minted}{clojure}
  (rep :a) ;=> IllegalArgumentException ...
\end{minted}

Methods are installed via {\clj{defmethod}}.
\begin{minted}{clojure}
(defmethod rep Keyword [x] (str (name x)))
\end{minted}
We extend \clj{rep}'s dispatch table, mapping
the \emph{dispatch value} \clj{Number} to the function
\clj{(fn [x] (str (name x)))}. The call \clj{(rep arg)}
uses the value of \clj{(isa? (disp-fn arg) disp-val)}
on each method entry to pick a winner \clj{mth}, and finally returns \clj{(mth arg)}.
\clj{isa?} is a subclassing check when provided with classes
\begin{minted}{clojure}
  (isa? Keyword Object) ;=> true
  (isa? Keyword Number) ;=> false
\end{minted}
otherwise an equality check.
\begin{minted}{clojure}
  (isa? :a :a) ;=> true
  (isa? :a 1) ;=> false
\end{minted}

For example
\clj{(rep :a)}
picks the \clj{Keyword} method because
\clj{(isa? (class :a) Keyword)} returns true,
and finally returns \clj{"a"}, the value of
\clj{((fn [x] (str (name x))) :a)}.
We give the full definition of \clj{rep}.

\begin{Code}
\begin{exmp}
\begin{minted}{clojure}
(ann rep [Any -> String])
(defmulti rep class)
(defmethod rep Keyword [x] (str (name x)))
(defmethod rep Number [x] (str (inc x)))

(rep :a) ;=> "a"
(rep 1)  ;=> "2"
\end{minted}
\end{exmp}
\end{Code}
Typed Clojure does not statically ensure multimethod calls dispatch successfully---
\clj{(rep "a")} type checks but throws a runtime error.

The flexibility of \clj{isa?} is key to the generality of multimethods. 
We can dispatch on the \clj{:op} key 
of our HMap AST \clj{Expr}.
Keywords are functions that look themselves up in their argument, so \clj{:op}
is our dispatch function.
\begin{Code}
\begin{exmp}
\inputminted[firstline=4,lastline=23]{clojure}{code/demo/src/demo/eg5.clj}
\end{exmp}
\end{Code}
The destructuring syntax \clj{(fn [{:keys [val] :as m}] ...)} binds
\clj{m} to the value of the first argument, and \clj{val} to \clj{(:val m)}.

\clj{isa?} is special with vectors---vectors of the
same length recursively call \clj{isa?} on the elements pairwise.
\begin{minted}{clojure}
  (isa? [Keyword :a] [Object :a]) ;=> true
  (isa? [Keyword Keyword] [Object Object]) ;=> true
  (isa? [Keyword Object] [Object Object]) ;=> false
\end{minted}

Now multiple dispatch is possible---we dispatch on the class of both
arguments simultaneously by defining a dispatch function that returns
a vector containing the classes of the arguments.
\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
\end{exmp}
\end{Code}
The dispatch values are also vectors---the first method is picked
when the left argument is a \clj{Number} and the right is a \clj{Keyword},
and is thus safe to increment and extract its name respectively.
The dispatch value \clj{:default} specifies a default
method if no preferred method is found.

No extra annotations are needed to follow type-directed control flow
in multimethod dispatch.

@@

The final example combines everything we will cover for the rest of the paper:
multimethod dispatch, reflection resolution via type hints, Java method
and constructor calls, conditional and exceptional flow reasoning
and use of HMaps. 
\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
\end{exmp}
\end{Code}
\clj{PayLoad} is either a HMap containing a file
or a string. We dispatch on \clj{:p} to distinguish the two cases---for example on \clj{:F}
we know the \clj{:file} is a file.
The body of the first method uses type-hints to resolve reflection
and conditional control flow to prove null-pointer exceptions are impossible.
The second method is similar except it uses exceptional control flow.
